Hydra Battles and Cie

### 2021-02-17 doc/chapter-primrec.tex
minor corrections

\emph{You may start this exercise with
\url{../exercises/primrec/FibonacciPR.v}.}

@@

\subsection{The Ackermann function}

Here is a possible definition, where (\texttt{iterate\,$f$\,$n$})
is the $n$-th iteration of $f$.

@@

@@

Please note that for any given $n$, the unary function
(\texttt{Ack\,$n$}) is primitive recursive.

@@

Then, our impossibility proof is just a sequence of easy small steps.

@@

  Fact Ack_majorized : majorized (Ack: naryFunc 2)  Ack.

  Section Contrad.
    Variable n: nat.
    Hypothesis  Hn: forall x y,  Ack x y <= Ack n (max x  y).

    Remark R01: n <> 0.

@@

    Remark R03: max 2 n = n.

    Remark R04: Ack n (2 * n + 3) = Ack n (Ack 2 n).

    Remark R05: Ack n (2 * n + 3) <= Ack (n + 2) n.

    Remark R06: Ack (n + 2) n <= Ack n (n + 2).

    Remark R07: Ack n (n + 2) < Ack n (2 * n + 3).

    Lemma contrad : False.

>>>>

\emph{You may start this exercise with the file
\url{../exercises/primrec/FibonacciPR.v}.}

@@

\subsection{Ackermann function}

Here is a possible definition, where (\texttt{iterate\,$f$\,$n$})
is the $n$-th iteration of $f$\,\footnote{Please not confuse with \texttt{primRec.iterate}, which is monomorphic and does not share the same order of arguments.}

@@

Another way of defining this function is 
using an inner \texttt{fix} (see for instance the discussion in \url{https://stackoverflow.com/questions/10292421/error-in-defining-ackermann-in-coq}).

We preferred to use the former definition, in order to
take advantage from generic lemmas about \texttt{iterate}.

Fixpoint Ack (n m : nat) : nat :=
  match n with
  | O => S m
  | S p => let fix Ackn (m : nat) :=
               match m with
               | O => Ack p 1
               | S q => Ack p (Ackn q)
               end
           in Ackn m
  end.

End Alt.
\end{Coqsrc}

@@

An important property of the Ackermann function helps us 
to overcome the difficulty raised by nested recursion, by climbing up the hierarchy $\texttt{Ack}\,n\,\_\;(n\in\mathbb{N})$.

\begin{Coqsrc}
Lemma Ack_Ack_le : forall k m n, 
    Ack k (Ack m n) <= Ack (2 + max k m) n.
\end{Coqsrc}

Please note also that for any given $n$, the unary function
(\texttt{Ack\,$n$}) is primitive recursive.

@@

Please note that, if the Ackermann function were primitive recursive, then there would exist some natural number $n$, such that, for all $x$ and $y$, the inequality 
$\texttt{Ack}\,x\,y\leq \texttt{Ack}\,n\,(\texttt{max}\,x\,y)$ holds.
Thus, our impossibility proof is just a sequence of easy small steps.

@@

  Fact Ack_majorized : majorized (Ack: naryFunc 2)  Ack.
 \end{Coqsrc}

 In order to share the variable \texttt{n}, let us open a section.

\begin{Coqsrc}
  Section Contrad.
    Variable n: nat.
    Hypothesis  Hn: forall x y,  Ack x y <= Ack n (max x  y).
\end{Coqsrc}

Firs, we get rid of the cases $n=0$ and $n=1$, thus we can rewrite
(\texttt{max 2 n}) into \texttt{n}.

  \begin{Coqsrc}
    Remark R01: n <> 0.

@@

    Remark R03: max 2 n = n.
\end{Coqsrc}

Please keep in mind that \texttt{(2 * n + 3)} can be safely replaced with 
\texttt{(Ack 2 n)}.

\begin{Coqsrc}
    Remark R04: Ack n (2 * n + 3) = Ack n (Ack 2 n).

    Remark R05: Ack n (2 * n + 3) <= Ack (n + 2) n.
\end{Coqsrc}

 By hypothesis \texttt{Hn}, the following inequality holds.

 \begin{Coqsrc}
    Remark R06: Ack (n + 2) n <= Ack n (n + 2).
 \end{Coqsrc}

By monotony of the Ackerman function, we get:

\begin{Coqsrc}
    Remark R07: Ack n (n + 2) < Ack n (2 * n + 3).
\end{Coqsrc}

which contradicts \texttt{R05}.

\begin{Coqsrc}
    Lemma contrad : False.
